---
title: "Assignment 2"
author: "Milan Filipovic"
date: "April 13, 2021"
output: html_document
---


```{r}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.align = 'center')
```

```{r, include = F, eval = T}
library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(faraway)
```

## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/16/2021

**Client Information**:

* Name: William Northway, David Edwards, Thomas Colby, and Wayne Dyer
*	Department: Stanford Medical Center
*	Phone: 650-723-5721
*	Name of principal investigator: William Northway, David Edwards, Thomas Colby, and Wayne Dyer, Stanford University

**Purpose**

* Statement of research hypothesis(es):
    + Factors associated with bronchopulmonary dysplasia (BPD) in newborn infants:
        1. Oxygen or intubation therapy cause BPD
        2. Pathological indices relate to BPD
        3. Background and health variables indicate high risk for contracting BPD
        4. BPD is serious to an infant’s life

*	Statement of statistical hypothesis(es):
	  1. Oxygen or intubation therapy cause BPD
        - Null hypothesis: Duration of oxygen or intubation therapy has no significant association with BPD
        - T-tests on log-transformed data for time length of oxygen or intubation therapy or logistic regression techniques can be used to assess association level with BPD. 
    2. Pathological indices relate to BPD
        - Null hypothesis: Bronchiola mucosa, pulmonary interstitium, hyalin membranes, alveolar infiltrate, inflation pattern, BPD stage based on lung tissue, hematoidin have no significant association with BPD
        - Chi-square tests can be used to understand the associations between each pathological index with BPD, since these are discrete ordinal variables.
    3. Background and health variables indicate high risk for contracting BPD
        - Null hypothesis: Age, sex, year of birth, one-minute APGAR score, estimated gestational age, birthweight, and age at onset of respiratory symptoms have no significant association with BPD
        -	Multiple logistic regression, t-tests or chi-square tests
    4. BPD is serious to an infant’s life
        - Null hypothesis: BPD is not significantly associated with survival or survival time
        - 2 proportion t-test given that we keep in mind differential follow-up times that might occur in this study sample.

**Background & references**:

* BPD is a respiratory problem that affects some newborn infants who have had respiratory distress syndrome (RDS) and subsequent oxygen therapy for it
* Oxygen or intubation therapy also suspected to cause BPD
* Severity of lung tissue deterioration measured on a scale of I to IV
* Stages I and II classified as not having BPD, stages III and IV classified as having BPD
* (Wolfe, “A Retrospective Study of Possible Causality Factors”)


**Descriptions**:

* Study design: Retrospective study

* Population(s): Newborn infants who have had RDS and subsequent oxygen therapy for it

* Sample(s): 299 infants treated at Stanford Medical center from 1962 – 1973 who had clinical and x-ray symptoms of RDS and received ventilatory assistance by intubation for over 24 hours

* Dependent variable(s):	
    + Radiographic BPD score
    + Survival
    + Survival time

* Independent variable(s):
    * Background/Health variables
        + Age
        + Sex
        + Year of birth
        + One-minute APGAR score
        + Estimated gestational age
        + Birthweight
        + Age at onset of respiratory symptoms
        + RDS severity score
    * Measures of the therapy
        + Exposure to low concentrations of elevated oxygen
        + Exposure to medium concentrations of elevated oxygen
        + Exposure to high concentrations of elevated oxygen
        + Duration of endotracheal intubation
        + Duration of assisted ventilation
    * Pathological indices (only available for infants that died)
        + Bronchiola mucosa
        + Pulmonary interstitium
        + Hyalin membranes
        + Alveolar infiltrate
        + Inflation pattern
        + BPD stage based on lung tissue
        + Hematoidin


## Question 2

*2.* Briefly outline statistical procedures that may help answer the four main statistical questions listed on Page 105 of Wolfe's casebook study.

**Solution**:

The four statistical questions listed on Page 105 are as follows:

1. Do oxygen or intubation therapy cause BPD?
    - Need to address that we cannot assess causation from a retrospective study and clarify that with the researchers. Analyses should include only infants that lived long enough to contract BPD (the 248 infants who survived 3+ days).
    - T-tests on log-transformed data for time length of oxygen or intubation therapy or logistic regression techniques can be used to assess association level with BPD. 
    
2. How are the pathological indices related to BPD?
    - Chi-square tests can be used to understand the associations between each pathological index with BPD, since these are discrete ordinal variables.

3. Which background or health variables indicates high risk for contracting BPD?
    -	Multiple logistic regression can be used to assess the associations between all continuous and discrete ordinal variables and outcome of contracting BPD. Additionally, we can use t-tests and chi-square tests for continuous and discrete ordinal/categorical variables, respectively.

4. How serious is BPD to an infant's life?
    - 2 proportion t-test given that we keep in mind differential follow-up times that might occur in this study sample.
    - Simple chi-square tests comparing BPD stage to survival (dead or alive).

Brief statistical procedures that may help answer the four main statistical questions include t-test, chi-square tests, and multiple regression techniques. Initial descriptive statistics must be conducted to examine the data and analyze the appropriate sample.


```{r read in dataset, echo = F}
getwd()
BPD = read_dta("bpd.dta")
BPD_tble <- as_tibble(BPD) %>% print()
```


```{r check variable type, echo = F}
glimpse(BPD_tble)
print(BPD_tble, width = Inf)

#create a categorical variable for gender
BPD_tble %>% mutate(gender = ifelse(BPD_tble$sex == 1, "male", "female")) 
```

## Question 3

*3a.* Create a demographic table describing the study population. (Who was studied? How many were studied? To whom could you generalize results?) Include the following descriptions: 


* % female; `gender`
* median and range of year of birth; `bxy`
* % APGAR scores; `agar`
* median and IQR gestational age; `gage`
* median and IQR birthweight; `bwt`
* median and IQR age at onset of respiratory symptoms; `rage`
* median and IQR age at onset of ventilatory assistance; `vage`

Demographic table had the study population split into with and without BPD, where with BPD included infants that reached at least stage III and without BPD included infants that did not reach stage III of disease progression.

```{r}
#
demographic_table <- BPD_tble %>% 
  mutate(gender = ifelse(BPD_tble$sex == 1, "male", "female"),
         bpdyesno = ifelse(bpd >= 30, "w/ BPD", "w/o BPD"),
         ) %>%  
  select(gender, bxy, agar, gage, bwt, rage, vage, bpdyesno) %>% 
  tbl_summary(
    by = bpdyesno,
    statistic = c("agar","gage","bwt", "rage", "vage") ~ "{median} ({p25}, {p75})",
    digits = all_continuous() ~ 1) %>% 
  add_overall() %>% 
  add_p(pvalue_fun = ~style_pvalue(.x, digits = 2)) %>% 
  bold_p() 
  
range(BPD_tble$bxy)
#range of birth year included in table attribute
demographic_table

```


*3b.* Plots (histograms or box plots) and summary statistics for the distributions of: 

i. hours of exposure to low concentrations of elevated oxygen 
ii. hours of exposure to medium concentrations of elevated oxygen 
iii. hours of exposure to high concentrations of elevated oxygen 
iv. hours of duration of endotracheal intubation 
v. hours of duration of assisted ventilation 


**Solution**

**i.** Hours of exposure to low concentrations of elevated oxygen 

```{r 3b i}

summary(BPD_tble$lo2)

BPD_tble %>% ggplot(mapping = aes(x = lo2)) +
  geom_histogram(binwidth = 200, fill="blue")+
  ggtitle("Hours of Exposure to Low Concentration (23-39%) of Elevated Oxygen") +
  labs(x = "Time (hours)") +
  theme_bw()

print(sort(BPD_tble$lo2))

```
We see that the distribution is right skewed, with outliers at the high extreme.


**ii.** Hours of exposure to medium concentrations of elevated oxygen

```{r 3b ii}

summary(BPD_tble$mo2)

BPD_tble %>% ggplot(mapping = aes(x = mo2)) +
  geom_histogram(fill="blue")+
  labs(x = "Time (hours)") +
  ggtitle("Hours of Exposure to Mid Concentration (40-79%) of Elevated Oxygen") +
  theme_bw()

sort(BPD_tble$mo2)

```

**iii.** Hours of exposure to high concentrations of elevated oxygen

```{r 3b iii}

summary(BPD_tble$ho2)

BPD_tble %>% ggplot(mapping = aes(x = ho2)) +
  geom_histogram(fill="blue") +
  labs(x = "Time (hours)") +
  ggtitle("Hours of Exposure to High Concentration (80-100%) of Elevated Oxygen") +
  theme_bw()

print(sort(BPD_tble$ho2))

```
This distribution is also right skewed. The scale of time is much lower then mid and low oxygen exposure.


**iv.** Hours of duration of endotracheal intubation

Next we examine the distribution of the duration of endotracheal intubation `idur`.
```{r 3b iv}

summary(BPD_tble$idur)

BPD_tble %>% ggplot(mapping = aes(x = idur)) +
  geom_histogram(fill = "blue") + 
  labs(x = "Time (hours)") +
  ggtitle("Duration of Endotracheal Intubation") +
  theme_bw()

print(sort(BPD_tble$idur))

```
The distribution appears heavily right skewed, with many potential outliers.

**v.** Hours of duration of assisted ventilation 

Next we examine the distribution of the duration of assisted ventilation intubation `vdur`.
```{r 3b v}

summary(BPD_tble$vdur)

BPD_tble %>% ggplot(mapping = aes(x = vdur)) +
  geom_histogram(fill = "blue") + 
  labs(x = "Time (hours)") +
  ggtitle("Duration of Ventillatory Assistance") +
  theme_bw()

print(sort(BPD_tble$vdur))

```

c. Counts (or bar graphs) of RDS and radiographic BPD scores. How many infants were classified as having RDS, very severe RDS, and BPD? 

Looking at the counts for RDS in our study population, we see that 296 (99.0%) infants were classified as having RDS (score of 1 or higher), of which 40 (13.4%) infants were classified as having very severe RDS (score of 5). 1 (0.0%) infant did not have any data on RDS classification. 

Looking at the counts for BPD in our study population, 71 (23.7%) infants were classified as having BPD (having reached stage III or stage IV). For the purposes of overall descriptive analysis here, the counts here did not filter for only infants that survived 3 days or more.


```{r}
# rds
BPD_tble %>% 
  mutate(rds = as.factor(rds)) %>% 
  ggplot(mapping = aes(rds)) +
  geom_bar(aes(fill = rds), colour = "black") +
  scale_fill_brewer(palette = "Pastel1") +
  labs(x = "RDS Score (0-5)", y = "Count",
       title = "RDS Severity Scores for Study Population (N = 299)") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3) +
  theme_bw()

# bpd
BPD_tble %>% 
  mutate(bpd = as.factor(bpd)) %>% 
  ggplot(mapping = aes(bpd)) +
  geom_bar(aes(fill = bpd), colour = "black") +
  scale_fill_brewer(palette = "Pastel1") +
  labs(x = "Stage of BPD (max radiological score x 10)", y = "Count",
       title = "Stages of BPD Present in Study Population (N = 299)") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3) +
  theme_bw()

```

d. Missing data and outliers. Identify variables with large numbers of missing 
data values (say, >20% missing). List variables with extreme outliers. 

Identifying variables with > 20% missing for a dataset containing 299 observations is equivalent to identifying variables with > 60 missing values.
We see that in the bpd data set, 8 variables (`agar`, `brmu`, `puin`, `hyme`. `alin`, `emph`, `lbpd`, `hema`) have more than 60 missing values.

```{r 3d, fig.height = 10, fig.width = 12}

# NA counts for each variable
missing <- BPD_tble %>%
  select(everything()) %>% 
  summarise_all(funs(sum(is.na(.)))) %>% 
  print(width = Inf)

# extreme outlier identification 
# excluded pid since pid is unique
# variables with 5 unique values or less were excluded 

df <- reshape2::melt(BPD_tble[,-c(1,2, 14, 18:24)])
ggplot(df, aes(x = value)) +
  facet_wrap(~variable, scales = "free_x") +
  geom_histogram() +
  theme_bw()

```


```{r}
par(mfrow = c(2,2))
halfnorm(BPD_tble$bxy, ylab = "Sorted Leverages")
title(main = "bxy")
halfnorm(BPD_tble$agar, ylab = "Sorted Leverages")
title(main = "agar")
halfnorm(BPD_tble$gage, ylab = "Sorted Leverages")
title(main = "gage")
halfnorm(BPD_tble$bwt, ylab = "Sorted Leverages")
title(main = "bwt")

halfnorm(BPD_tble$rage, ylab = "Sorted Leverages")
title(main = "rage")
halfnorm(BPD_tble$vage, ylab = "Sorted Leverages")
title(main = "vage")
halfnorm(BPD_tble$idur, ylab = "Sorted Leverages")
title(main = "idur")
halfnorm(BPD_tble$vdur, ylab = "Sorted Leverages")
title(main = "vdur")

halfnorm(BPD_tble$lo2, ylab = "Sorted Leverages")
title(main = "lo2")
halfnorm(BPD_tble$mo2, ylab = "Sorted Leverages")
title(main = "mo2")
halfnorm(BPD_tble$ho2, ylab = "Sorted Leverages")
title(main = "ho2")
halfnorm(BPD_tble$surv, ylab = "Sorted Leverages")
title(main = "surv")


```
```{r, echo = F}
print("Influential Observations in agar")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(36, 110))

print("Influential Observations in bwt")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(48, 253)) 

print("Influential Observations in rage")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(34, 225)) 

print("Influential Observations in vage")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(98, 186)) 

print("Influential Observations in idur and vdur")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(173, 260)) 

print("Influential Observations in lo2")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(168, 212)) 

print("Influential Observations in mo2")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(20, 209)) 

print("Influential Observations in ho2")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(20, 63)) 

print("Influential Observations in surv")
BPD_tble %>% 
  select(-c(2, 14)) %>% 
  slice(c(5, 1)) 


```

