---
title: 'Assignment 3: Progression of Coronary Artery Calcification in Normal Subjects'
author: "Milan Filipovic"
date: "April 21, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F, eval = T}
library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(purrr)
```

**Exercise**

Complete a Biostatistics Consulting Worksheet for the following study. Your worksheet 
should include: 

* Statement of statistical hypothesis(es) 
* Assessment of assumptions used in any statistical inference 
* Analysis of data set. (You will need to retrieve from our website one of two data 
  files: “cac.dta” and “cac.txt”.) 
* Interpretation of results in terms of original research hypothesis so that a non-statistician would be able to explain results to colleagues.


## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/23/2021

**Client Information**:

* Name: H.C. Yoon, A. Emerick, J Hill and J Goldin
*	Department: Department of Diagnostic Imaging, Kaiser Moanalua Medical Center and the Department of Radiological Sciences, UCLA
*	Phone: (310) 267-8785
*	Name of principal investigator: H.C. Yoon, A. Emerick, J Hill and J Goldin from the Department of Diagnostic Imaging, Kaiser Moanalua Medical Center and the Department of Radiological Sciences, UCLA

**Purpose**

* Statement of research hypothesis(es):
    + To test the hypothesis that the rate of CAC progression is gender specific being greater in men then women.

*	Statement of statistical hypothesis(es):
    + The rate of CAC progression in men is greater than it is in women
    + Null hypothesis: The rate of change in calcium volume score (CVS) between men and women is statistically significantly different

**Background & references**:

* Recent reports indicate that electron-beam computed tomography (EBCT) can document 
the presence and monitor the progression of atherosclerotic CAC in the general adult population as well 
as in those with increased cardiovascular risk (1-3). 
* EBCT can measure changes in the extent of CAC in adults treated with lipid-lowering agents.
*Interpretation of the clinical significance of different coronary artery calcium scores in the same patient is dependent on several factors:
    + Measurement variation
    + Expected rate of progression of coronary calcium
    
1. Maher JE; Bielak LF; Raz JA; Sheedy II PF; Schwartz RS; Peyser PA. Progression of Coronary Artery Calcification: A Pilot Study. Mayo Clin Proc 1999; 74:347-355. 1999.
2. Goodman WG, Goldin J, Kuizon BD, Yoon C, Gales B, Sider D, Wang Y, Chung J, Emerick A, Greaser L, Elashoff RM, Salusky IB. Coronary-artery calcification in young adults with end-stage renal disease who are undergoing dialysis. N Engl J Med. 2000;342:1478-83.
3. Budoff MJ; Lane KL; Bakhsheshi H; Mao S; Grassmann BO; Friedman BC; Brundage BH. Rates of Progression of Coronary Calcium by Electron Beam Tomography. Am J Cardiol 2000; 86:8-11.


**Descriptions**:

* Study design: Retrospective study

* Population(s): Adults who experience the progression of athersclerotic CAC

* Sample(s): 217 asymptomatic subjects who underwent at least two electron-beam computed tomography (EBCT) for detection of CAC as a part of a clinical screening program. 
    + Asymptomatic defined as no history of ischemic heart disease 
    + No abnormal electrocardiogram, stress test, coronary angiogram, and no prior myocardial infarction or coronary bypass surgery

* Dependent variable(s):	
    + vol1
    + vol2
    + days

* Independent variable(s):
    + sex
    + age



```{r read/save dataset, eval = F, echo = F}
cac = read_dta("cac.dta")
cac_tble <- as_tibble(cac)
fwrite(cac_tble, file = "CAC_data")
```


```{r loading tibble, echo = FALSE}
cac_tble <- fread("CAC_data") %>% print()
```


Checking Data for cleanliness 

Variable Descriptions
```{r variable info}
#Haven stores the stata labels in as variable attributes. We can access them using map_chr from the purrr package
#cac %>% map_chr(~attributes(.)$label)
```

Note about CAC score taken from radiopaedia:
CAC score of 1-112: low risk with a relative risk ratio of 1.9 (95 CI: 1.3-2.8) 
CAC score of 100-400: moderate risk with a relative risk ratio of 4.3 (95% CI:3.1-6.1) 
CAC score of 401-999: high risk with a relative risk ratio of 7.2 (95% CI:5.2-9.9)
Source: [link](https://radiopaedia.org/articles/calcium-volume-score?lang=us#:~:text=CAC%20score%20of%201%2D112,%25%20CI%3A5.2%2D9.9))


```{r}
cac_tble %>% mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1) %>%
  select(-sex, -nid) %>%
  tbl_summary(by = gender, missing_text = "Missing",
              label = list(age ~ "Age First Scanned",
                           vol1 ~ "CVS at First Visit",
                           vol2 ~ "CVS at Second Visit",
                           days ~ "Days Between Visits",
                           vol_change ~ "Change in CVS")) %>% 
  add_p(test = list(all_continuous() ~ "t.test")) %>% 
  bold_labels() %>% 
  add_overall() %>% 
  modify_header(label ~ "**Participant Characteristics**")
```




Next Visualize key predictors 
```{r}
cac_tble <- cac_tble %>%
  mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1)

summary(cac_tble)

cac_tble %>% ggplot(mapping = aes(x = age)) +
  geom_histogram(mapping = aes(x = age, fill = gender)) +
  labs(title = "Age when first scanned")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = vol_change, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Change in Calcium Volume Score")
```


Age appears to be normally distributed, wheareas volume change appears to be right skewed. Some transformation should be applied to normalize the data, and outliers should be examined for validity. 


