---
title: 'Assignment 3: Progression of Coronary Artery Calcification in Normal Subjects'
author: "Milan Filipovic, Vivian Truong, Lillian Chen"
date: "April 21, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F, eval = T}

library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(purrr)
library(sjlabelled)

```

**Exercise**

Complete a Biostatistics Consulting Worksheet for the following study. Your worksheet 
should include: 
* Statement of statistical hypothesis(es) 
*  Assessment of assumptions used in any statistical inference 
* Analysis of data set. (You will need to retrieve from our website one of two data 
  files: “cac.dta” and “cac.txt”.) 
*Interpretation of results in terms of original research hypothesis so that a non-statistician would be able to explain results to colleagues.


## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/23/2021

**Client Information**:

* Name: H.C. Yoon, A. Emerick, J Hill and J Goldin
*	Department: Department of Diagnostic Imaging, Kaiser Moanalua Medical Center and the Department of Radiological Sciences, UCLA
*	Phone: (310) 267-8785
*	Name of principal investigator: H.C. Yoon, A. Emerick, J Hill and J Goldin from the Department of Diagnostic Imaging, Kaiser Moanalua Medical Center and the Department of Radiological Sciences, UCLA

**Purpose**

* Statement of research hypothesis(es):
    + To test the hypothesis that the rate of CAC progression is gender specific being greater in men then women.

*	Statement of statistical hypothesis(es):
    + The rate of CAC progression in men is greater than it is in women
    + Null hypothesis: The rate of change in calcium volume score (CVS) between men and women is statistically significantly different

**Background & references**:

* Recent reports indicate that electron-beam computed tomography (EBCT) can document 
the presence and monitor the progression of atherosclerotic CAC in the general adult population as well 
as in those with increased cardiovascular risk (1-3). 
* EBCT can measure changes in the extent of CAC in adults treated with lipid-lowering agents.
*Interpretation of the clinical significance of different coronary artery calcium scores in the same patient is dependent on several factors:
    + Measurement variation
    + Expected rate of progression of coronary calcium


**Descriptions**:

* Study design: Retrospective study

* Population(s): Adults who experience the progression of athersclerotic CAC

* Sample(s): 217 asymptomatic subjects who underwent at least two electron-beam computed tomography (EBCT) for detection of CAC as a part of a clinical screening program. 
    + Asymptomatic defined as no history of ischemic heart disease 
    + No abnormal electrocardiogram, stress test, coronary angiogram, and no prior myocardial infarction or coronary bypass surgery

* Dependent variable(s):	
    + vol1
    + vol2
    + days

* Independent variable(s):
    + sex
    + age



```{r read save and load dataset, echo = F}

cac <- read_dta("cac.dta") %>% as_tibble()

fwrite(cac, file = "CAC_data")

cac_tble <- fread("CAC_data") %>% print()

```


Checking Data for cleanliness 

Variable Descriptions
```{r variable info, eval = F}
# Haven stores the stata labels in as variable attributes. We can access them 
# using map_chr from the purrr package
(varlabels <- cac_tble %>% map_chr(~attributes(.)$label))

```

```{r variable info base R, eval = F}

# Extracting variable data using base R functions
(varlabels <- str(lapply(cac_tble, attr, "label")))

```
The variable labels are as follows:

 * nid : "id subject number, 1-217"
 * sex : "rf male subject, 0n/1y"
 * age : "rf age when first scanned, yrs"
 * vol1: "ct visit #1 CVS, mm3"
 * vol2: "ct visit #2 CVS, mm3"
 * days: "ct elapsed days between visits"


Note about CAC score taken from radiopaedia:
CAC score of 1-112: low risk with a relative risk ratio of 1.9 (95 CI: 1.3-2.8) 
CAC score of 100-400: moderate risk with a relative risk ratio of 4.3 (95% CI:3.1-6.1) 
CAC score of 401-999: high risk with a relative risk ratio of 7.2 (95% CI:5.2-9.9)
Source: [link](https://radiopaedia.org/articles/calcium-volume-score?lang=us#:~:text=CAC%20score%20of%201%2D112,%25%20CI%3A5.2%2D9.9))

```{r table 1 participant characteristics}
cac_tble %>% mutate(gender = ifelse(cac_tble$sex == 1, "Men","Women"), 
                    vol_change = vol2-vol1,
                    ratechange = vol_change/days) %>%
  select(-sex, -nid) %>%
  tbl_summary(by = gender, missing_text = "Missing",
              label = list(age ~ "Age First Scanned",
                           vol1 ~ "CVS at First Visit",
                           vol2 ~ "CVS at Second Visit",
                           days ~ "Days Between Visits",
                           vol_change ~ "Change in CVS",
                           ratechange ~ "Rate of Change in CVS")) %>% 
  add_p(test = list(all_continuous() ~ "t.test")) %>% 
  bold_labels() %>% 
  bold_p() %>% 
  add_overall() %>% 
  modify_header(label ~ "**Participant Characteristics**")
```




Next Visualize key predictors 
```{r}
cac_tble <- cac_tble %>%
  mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), 
         vol_change = vol2-vol1,
         ratechange = vol_change / days)

summary(cac_tble)

cac_tble %>% ggplot(mapping = aes(x = age)) +
  geom_histogram(mapping = aes(x = age, fill = gender)) +
  labs(title = "Age at First CT Scan")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = vol_change, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Change in Calcium Volume Score")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = ratechange, fill = gender)) +
  labs(x = expression(CAC~rate~(mm^3/day)), title = "Rate of Change in Calcium Volume Score")
```


Age appears to be normally distributed, volume change appears to be right skewed, and rate of volume change is also right skewed. Some transformation should be applied to normalize the data, and outliers should be examined for validity. Additionally, the values for volume change and rate of volume change have a large number of 0s, so our data is zero-inflated and should be noted so that log transformations include an offset.


```{r log transforms and visualizations}

cac_tble <- cac_tble %>%
  mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), 
         vol_change = vol2-vol1,
         ratechange = vol_change / days,
         logvol1 = log(vol1 + 1),
         logvol2 = log(vol2 + 1),
         logdiffvol_change = logvol2 - logvol1,
         logratechange = log(ratechange + 1))

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = logdiffvol_change, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Change in Log Calcium Volume Score")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = logratechange, fill = gender)) +
  labs(x = expression(CAC~rate~(mm^3/day)), title = "Log Rate of Change in Calcium Volume Score")

cac_tble %>% ggplot() +
  geom_histogram(mapping = aes(x = logdiffvol_change, fill = gender)) +
  labs(x = expression(log~CAC~(mm^3)), title = "Change in Log Calcium Volume Score")

cac_tble %>% ggplot() +
  geom_histogram(mapping = aes(x = logratechange, fill = gender)) +
  labs(x = expression(log~CAC~rate~(mm^3/day)), title = "Log Rate of Change in Calcium Volume Score")

```

The data is still extremely zero-inflated. It seems like the best course of action is not a linear regression, since despite transformations the distribution of the outcome is not normally distributed. Since there is no transformation that can be applied for us to meet normality assumptions of parametric tests, a possible route to take is to do a nonparametric test on the rate of volume change between men and women.


The Mann-Whitney U Test may be able to test the null hypothesis $H_0$: There is no difference in the rate of volume change between men and women in the study population.
```{r nonparametric test}

cac_men <- cac_tble %>% 
  filter(gender == "men")
cac_women <- cac_tble %>% 
  filter(gender == "women")

# Mann-Whitney U test - the statements below yield the same result 
# but are different ways of writing it
wilcox.test(cac_men$ratechange, cac_women$ratechange, paired = F)
wilcox.test(cac_tble$ratechange~cac_tble$gender)

median(cac_men$ratechange)
median(cac_women$ratechange)
```
The Mann-Whitney U test showed that there was a significant difference (W = 7092, p = .007) in the rate of volume change between men and women in the study population. The median rate of volume change for men was 0.020 mm^3^/day, and the median rate of volume change for women was 0 mm^3^/day.
