---
title: 'Assignment 2: Progression of Coronary Artery Calcification in Normal Subjects'
author: "Milan Filipovic"
date: "April 21, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F, eval = T}
library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(purrr)
```


## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/16/2021

**Client Information**:

* Name: 
*	Department: 
*	Phone: 
*	Name of principal investigator: 

**Purpose**

* Statement of research hypothesis(es):
    + To test the hypothesis that the rate of CAC progression is gender specific being greater in men then women.

*	Statement of statistical hypothesis(es):


**Background & references**:

* Recent reports indicate that electron-beam computed tomography (EBCT) can document 
the presence and monitor the progression of atherosclerotic CAC in the general adult population as well 
as in those with increased cardiovascular risk (1-3). 
* EBCT can measure changes in the extent of CAC in adults treated with lipid-lowering agents.
*Interpretation of the clinical significance of different coronary artery calcium scores in the same patient is dependent on several factors:
    + Measurement variation
    + Expected rate of progression of coronary calcium


**Descriptions**:

* Study design: 

* Population(s): 

* Sample(s): 217 Asymptomatic Subjects who underwent at least two electron-beam computed tomography (EBCT) for detection of CAC as a part of a clinical screening program. 
    +Asymptomatic defined as no history of ischemic heart disease 
    +No abnormal electrocardiogram, stress test, coronary angiogram, and no prior myocardial infarction or coronary bypass surgery

* Dependent variable(s):	
    + 

* Independent variable(s):
    +

**Exercise**

Complete a Biostatistics Consulting Worksheet for the following study. Your worksheet 
should include: 
* Statement of statistical hypothesis(es) 
*  Assessment of assumptions used in any statistical inference 
* Analysis of data set. (You will need to retrieve from our website one of two data 
  files: “cac.dta” and “cac.txt”.) 
*Interpretation of results in terms of original research hypothesis so that a nonstatistician would be able to explain results to colleagues.



```{r read/save dataset, eval = F, echo = F}
cac = read_dta("cac.dta")
cac_tble <- as_tibble(cac)
fwrite(cac_tble, file = "CAC_data")
```


```{r loading tibble, echo = FALSE}
cac_tble <- fread("CAC_data") %>% print()
```


Checking Data for cleanliness 

Variable Descriptions
```{r variable info}
#Haven stores the stata labels in as variable attributes. We can access them using map_chr from the purrr package
cac %>% map_chr(~attributes(.)$label)
```

```{r}
cac_tble %>% mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1) %>%
  select(-sex, -nid) %>%
  tbl_summary(by = gender, missing_text = "Missing",
              label = list(age ~ "Age First Scanned",
                           vol1 ~ "CVS at First Visit",
                           vol2 ~ "CVS at Second Visit",
                           days ~ "Days Between Visits",
                           vol_change ~ "Change in CVS")) %>% 
  bold_labels() %>% 
  add_overall() %>% 
  modify_header(label ~ "**Participant Characteristics**")
```

Next Visualize key predictors 
```{r}
cac_tble <- cac_tble %>%
  mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1)

summary(cac_tble)

cac_tble %>% ggplot(mapping = aes(x = age)) +
  geom_histogram(fill="lightblue") +
  labs(title = "Age when first scanned")

cac_tble %>% ggplot(mapping = aes(x = vol_change)) + geom_histogram()
```


Age appears to be normally distributed, wheareas volume change appears to be right skewed. Some transformation should be applied to normalize the data, and outliers should  be examined for validity. 




