---
title: 'Assignment 2: Progression of Coronary Artery Calcification in Normal Subjects'
author: "Milan Filipovic"
date: "April 21, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F, eval = T}
library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(faraway)
```


## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/16/2021

**Client Information**:

* Name: 
*	Department: 
*	Phone: 
*	Name of principal investigator: 

**Purpose**

* Statement of research hypothesis(es):
    + To test the hypothesis that the rate of CAC progression is gender specific being greater in men then women.

*	Statement of statistical hypothesis(es):


**Background & references**:

* Recent reports indicate that electron-beam computed tomography (EBCT) can document 
the presence and monitor the progression of atherosclerotic CAC in the general adult population as well 
as in those with increased cardiovascular risk (1-3). 
* EBCT can measure changes in the extent of CAC in adults treated with lipid-lowering agents.
*Interpretation of the clinical significance of different coronary artery calcium scores in the same patient is dependent on several factors:
    + Measurement variation
    + Expected rate of progression of coronary calcium


**Descriptions**:

* Study design: 

* Population(s): 

* Sample(s): 

* Dependent variable(s):	
    + 

* Independent variable(s):
    +

**Exercise**

Complete a Biostatistics Consulting Worksheet for the following study. Your worksheet 
should include: 
* Statement of statistical hypothesis(es) 
*  Assessment of assumptions used in any statistical inference 
* Analysis of data set. (You will need to retrieve from our website one of two data 
  files: “cac.dta” and “cac.txt”.) 
*Interpretation of results in terms of original research hypothesis so that a nonstatistician would be able to explain results to colleagues.



```{r read in dataset, echo = F}
setwd("C:/Graduate School/Coursework/Spring 2021/402A/R projects/402A-spring2021/Assignment 3")
cac = read_dta("cac.dta")
cac_tble <- as_tibble(BPD) %>% print()
```
