---
title: 'Assignment 2: Progression of Coronary Artery Calcification in Normal Subjects'
author: "Milan Filipovic"
date: "April 21, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = F, eval = T}
library(haven)
library(tidyverse)
library(data.table)
library(gtsummary)
library(ggplot2)
library(purrr)
```


## Question 1

*1.* Based on the introduction and variable list, complete the following sections of a Biostatistics Consulting Worksheet:

a. Date & Client Information
b. Purpose
c. Background & references
d. Descriptions

**Date**: 04/16/2021

**Client Information**:

* Name: 
*	Department: 
*	Phone: 
*	Name of principal investigator: 

**Purpose**

* Statement of research hypothesis(es):
    + To test the hypothesis that the rate of CAC progression is gender specific being greater in men then women.

*	Statement of statistical hypothesis(es):


**Background & references**:

* Recent reports indicate that electron-beam computed tomography (EBCT) can document 
the presence and monitor the progression of atherosclerotic CAC in the general adult population as well 
as in those with increased cardiovascular risk (1-3). 
* EBCT can measure changes in the extent of CAC in adults treated with lipid-lowering agents.
*Interpretation of the clinical significance of different coronary artery calcium scores in the same patient is dependent on several factors:
    + Measurement variation
    + Expected rate of progression of coronary calcium


**Descriptions**:

* Study design: 

* Population(s): 

* Sample(s): 217 Asymptomatic Subjects who underwent at least two electron-beam computed tomography (EBCT) for detection of CAC as a part of a clinical screening program. 
    +Asymptomatic defined as no history of ischemic heart disease 
    +No abnormal electrocardiogram, stress test, coronary angiogram, and no prior myocardial infarction or coronary bypass surgery

* Dependent variable(s):	
    + 

* Independent variable(s):
    +

**Exercise**

Complete a Biostatistics Consulting Worksheet for the following study. Your worksheet 
should include: 
* Statement of statistical hypothesis(es) 
*  Assessment of assumptions used in any statistical inference 
* Analysis of data set. (You will need to retrieve from our website one of two data 
  files: “cac.dta” and “cac.txt”.) 
*Interpretation of results in terms of original research hypothesis so that a nonstatistician would be able to explain results to colleagues.



```{r read/save dataset, eval = F, echo = F}
cac = read_dta("cac.dta")
cac_tble <- as_tibble(cac)
fwrite(cac_tble, file = "CAC_data")
```


```{r loading tibble, echo = FALSE}
cac_tble <- fread("CAC_data") %>% print()
```


Checking Data for cleanliness 

Variable Descriptions
```{r variable info}
#Haven stores the stata labels in as variable attributes. We can access them using map_chr from the purrr package
cac %>% map_chr(~attributes(.)$label)
```

Note about CAC score taken from radiopaedia:
CAC score of 1-112: low risk with a relative risk ratio of 1.9 (95 CI: 1.3-2.8) 
CAC score of 100-400: moderate risk with a relative risk ratio of 4.3 (95% CI:3.1-6.1) 
CAC score of 401-999: high risk with a relative risk ratio of 7.2 (95% CI:5.2-9.9)
Source: [link](https://radiopaedia.org/articles/calcium-volume-score?lang=us#:~:text=CAC%20score%20of%201%2D112,%25%20CI%3A5.2%2D9.9))


```{r}
#change in score would not usually be added toa demographic table but I do it here
cac_tble %>% mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1) %>%
  select(-sex, -nid) %>%
  tbl_summary(by = gender, missing_text = "Missing",
              label = list(age ~ "Age First Scanned",
                           vol1 ~ "CVS at First Visit",
                           vol2 ~ "CVS at Second Visit",
                           days ~ "Days Between Visits",
                           vol_change ~ "Change in CVS")) %>% 
  add_p(test = list(all_continuous() ~ "t.test")) %>% 
  bold_labels() %>% 
  add_overall() %>% 
  modify_header(label ~ "**Participant Characteristics**")
```




Next Visualize key predictors 
```{r}
cac_tble <- cac_tble %>%
  mutate(gender = ifelse(cac_tble$sex == 1, "men","women"), vol_change = vol2-vol1)

summary(cac_tble)

cac_tble %>% ggplot(mapping = aes(x = age)) +
  geom_histogram(mapping = aes(x = age, fill = gender)) +
  labs(title = "Age when first scanned")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = vol_change, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Change in Calcium Volume Score")


cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = vol1, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Initial Calcium Volume Score")

cac_tble %>% ggplot() +
  geom_boxplot(mapping = aes(x = vol2, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Follow-Up Calcium Volume Score")
```


Age appears to be normally distributed, wheareas CAC volume variables all appear right skewed. Some transformation should be applied to normalize the data, and outliers should be examined for validity. 



```{r}
log_cac<- cac_tble %>% mutate(logvol1 = log(vol1 +1), logvol2 = log(vol2 +1)) %>% 
  mutate(logvol_change = logvol2 - logvol1)


log_cac %>% ggplot() +
  geom_histogram(mapping = aes(x = logvol1, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Log-Tranformed Initial CVS")

log_cac %>% ggplot() +
  geom_histogram(mapping = aes(x = logvol2, fill = gender)) +
  labs(x = expression(CAC~(mm^3)), title = "Log-Tranformed Follow-up CVS")
```

After a shifted log transformation, the the distribution for log CAC scores at both visits are still highly zero-inflated, although skewness has been reduced. 

```{r}
mod <- glm(logvol2 ~ age + days + logvol1 + sex, data = log_cac) 
mod %>% tbl_regression()
```


Instead a two sample wilcoxon rank sum test is run on the two gender with cac score change to test for differences. A two sample welch's test was performed previously, but due to non-normality of data, a nonparametric test will be conducted between the two genders. 
```{r}
wilcox.test(vol_change ~ gender, paired = FALSE, data = cac_tble)
```

The wilcoxon rank sum test determined that there is a significant difference in CVS score change between men and women. 
For future analysis, it may be easier to analyze if categorical ranking system is implemented for risk determination, or alternate scores were used to assess outcome. 




