---
title: "B402A Assignment 7"
author: "Lillian Chen, Milan Filipovic, Vivian Truong"
date: "5/24/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(haven)
library(gt)
library(tidyr)
library(ggforce) #can be used for large facet grids
```

```{r}

data <- readxl::read_xlsx("./20210521_402a.xlsx", col_names = T, skip = 1)

str(data)

#lalalalalalalalala

dataclean <- data %>% 
  rename(sample = "Sample #",
         patientid = "Patient ID",
         IRI = "IRI Score (–,+)",
         visittype = "Visit Type") %>% 
  mutate(across(starts_with("CYTO"),
                ~ dplyr::case_when(
                  str_starts(.x, "<") == T ~ as.numeric(str_sub(.x, 2, -2)),
                  str_starts(.x, ">") == T ~ as.numeric(str_sub(.x, 2, -2)) + 1,
                  str_starts(.x, "1|2|3|4|5|6|7|8|9|0") == T ~ as.numeric(.x)
                )))

```

```{r}
library(data.table)
isidata <- fread(file = "dataclean.csv", header = T, stringsAsFactors = F)
(isi_tble <- tibble(isidata))
```

We begin by visualizing the distribution of cytokine values 



<<<<<<< HEAD


=======
```{r}


isi_tble %>% 
  select(CYTO1:CYTO6) %>% 
  gather() %>% 
  ggplot(aes(value)) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram()

isi_tble %>% 
  select(CYTO7:CYTO12) %>% 
  gather() %>% 
  ggplot(aes(value)) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram()

isi_tble %>% 
  select(CYTO13:CYTO18) %>% 
  gather() %>% 
  ggplot(aes(value)) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram()

isi_tble %>% 
  select(CYTO19:CYTO23) %>% 
  gather() %>% 
  ggplot(aes(value)) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram()

## my attempt to facet in one plot. Does not look good
isi_tble %>% select(starts_with("CYT")) %>%
  gather() %>%
  ggplot(aes(value)) +
facet_wrap(~ key, scales ="free") +
# facet_grid_paginate(~ key, ncol = 3, nrow = 3, page = 2)
  geom_boxplot()
```
>>>>>>> 2ed165577afd5beda6f141f540a003e5c4b7ce24

The Cytokine distributions are characterized by heavy right skew. 

<<<<<<< HEAD
```

=======
Now log transform all Cytokines
```{r}
>>>>>>> 2ed165577afd5beda6f141f540a003e5c4b7ce24

isi_tble %>% 
  select(CYTO1:CYTO6) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO7:CYTO12) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO13:CYTO18) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO19:CYTO23) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_histogram() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")


```

<<<<<<< HEAD

```{r Research Question 2}

xtabs(~ Inflammation, data = dataclean)
xtabs(~ Necrosis, data = dataclean)

# split outcomes (Inflammation or Necrosis Score) with 0-1 vs 2+ as -/+
# try two types of models: ordered factor outcome and binary outcome (try both)
=======
```{r}

isi_tble %>% 
  select(CYTO1:CYTO6) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_boxplot() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO7:CYTO12) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_boxplot() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO13:CYTO18) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_boxplot() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")

isi_tble %>% 
  select(CYTO19:CYTO23) %>% 
  gather() %>% 
  ggplot(aes(log(value))) + 
 facet_wrap(~ key, nrow = 3, scales = "free") + 
  geom_boxplot() + 
  labs(title = "Log Transformed") +
  xlab("log pg/mL")


```

* Among the log transformed cytokine values, it is seen that that overall normality is improved. 
 + CYTO12 and CYTO 19 appears to have a similar distribution when looking at the histogram, with large left skew and lack of density left of the peak density. 
 + CYTO13 appears as if it may have another peak in the right tail. It is right skewed. 
* CYTO 15 appears to have several outliers in the higher range.
* CYTO16 appears right skewed. 
* CYTO18 appears to have lower outliers. 
* CYTO17 has many values below level of detection, therefore the transformation did not affect overall normality. 
* CYTO21 appears to have a large right skew, though it should be noted that many values were below level of detection in CYTO 21. 


>>>>>>> 2ed165577afd5beda6f141f540a003e5c4b7ce24



